A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-06-04 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: minor changes

In this paper, we will characterize a generic space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy function) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurements (as in the number of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

@@

Since ensemble spaces are a convex subset of a locally convex topological vector space, they can be connected to Choquet theory. First of all, let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $\mu : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_X f d\mu$ for every continuous linear functional $f$ on $V$. The Choquet theorem states:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}

First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows to think go $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

We should now look at the requirements of the theorem, to see whether we can always apply it to an ensemble space. First of all, and ensemble space $\Ens$ is a convex subset of a convex sapce $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measure over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore the theorem does not apply in general.

We can apply Choquet theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allows a representation through a probability on points that may corresponds idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

@@

Another way to represent quantum ensembles is through the use of quasiprobability distributions. Given a space $X$, a measure $\mu$ is affine if $\mu(X)$. All probability measures are affine measures, but the reverse is not true as affine measures can be negative on particular regions. The Wigner function, for example, can be negative over small regions of phase space.

@@

>>>>

In this paper, we will characterize a generic space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy function) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurements (as in the dimension of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

@@

[This section has to make clear in what extra conditions on ensembles spaces Choquet theory can be applied.]

% Intro paragraph that says quickly how to connect to Choquet theory and its limitations. It should be a quick summary of the whole section
Since ensemble spaces are a convex subset of a locally convex topological vector space, they can be connected to Choquet theory.

% Quick review of Choquet theory
Let us first give a summary of the main results in Choquet theory that are interesting for our purposes. First of all, let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $\mu : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_E f d\mu$ for every continuous linear functional $f$ on $V$. The Choquet theorem states:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}

% Physical meaning of representation with a measure.
First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows to think go $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

% Limitation of applying Choquet theory to our case
We should now look at the requirements of the theorem, to see whether we can always apply it to an ensemble space. First of all, and ensemble space $\Ens$ is a convex subset of a convex sapce $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measure over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore the theorem does not apply in general.

% Define "Choquet space" as those spaces where we can apply Choquet theory
We can apply Choquet theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allows a representation through a probability on points that may corresponds idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

@@

Another way to represent quantum ensembles is through the use of quasiprobability distributions. Given a topological space $X$, a measure $\mu$ is affine if $\mu(X)=1$. All probability measures are affine measures, but the reverse is not true as affine measures can be negative on particular regions. The Wigner function, for example, can be negative over small regions of phase space.

@@

TODO: disjoint supports of Wigner functions do not give orthogonal subspaces (CHECK)
